Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic.

### 2024-12-16 paper.tex
change to quarto + add ola 7

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
%
\documentclass[
  12pt,
  a4paper,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic},
%   colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{times}
\hypersetup {colorlinks = true, linkcolor = blue, urlcolor = blue}
\usepackage{caption}
\captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
\usepackage{graphicx}
\usepackage{float}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic}
    \usepackage[]{authblk}
    \setlength{\affilsep}{2em}
    \renewcommand*{\Affilfont}{\normalsize\normalfont}
    \renewcommand\Authsep{, }
    \renewcommand\Authand{ \& }
    \renewcommand\Authands{, \& }
                                        \author[1, 3]{Juan Carlos Castillo}
                                                            \affil[]{Department of Sociology, Universidad de Chile \thanks{\href{mailto:juancastillov@uchile.cl}{\nolinkurl{juancastillov@uchile.cl}}}}
                                                                                \author[2]{Julio Iturra}
                                                            \affil[]{Bremen International Graduate School of Social Sciences, Universität Bremen}
                                                                                \author[3]{Kevin Carrasco}
                                                            \affil[]{Center for Social Conflict and Cohesion Studies}
                                            \date{}

\providecommand{\keywords}[1]
{
  \small
  \textbf{Keywords:} #1
}

\begin{document}
\maketitle
\begin{abstract}
Education is considered a key tool for social mobility and equality of opportunities. However, despite the widespread value of education, disparities in educational outcomes persist over time. To what extent are such educational disparities justified in society? And, what are the main factors driving this kind of justification? This research focuses on the changes in the justification of inequality in access to education in Chile and their links with the perception of meritocracy. The Chilean case offers an interesting context for this study given its high economic inequality, deep neoliberal policies, and commodification of social services such as education. The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic), which could be linked to a challenge of meritocratic ideals. For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 -- 2022, 6 waves, N = 2,927). Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education. The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.

.
\newline
\newline

@@

\setstretch{1.5}
\pagebreak

@@

Despite the widespread recognition of education as a key factor for social mobility and equality of opportunities, disparities in this domain are far from being eliminated. As the last PISA Report points out, ``Socio-economically disadvantaged students are seven times more likely than advantaged students to score below Level 2 in mathematics on average across OECD countries'' (\citeproc{ref-oecd_pisa_2023}{OECD, 2023}). Furthermore, empirical evidence from social science research has revealed an exacerbation of educational inequalities across various domains, such as family socioeconomic status, gender, ethnicity, and migratory background, particularly after COVID-19 (\citeproc{ref-darmodyImpactsCOVID19Control2021}{Darmody et al., 2021}). Nevertheless, and despite the consistent empirical evidence on educational inequalities and their consequences, research on distributive justice indicates remarkable individual differences in the justification of unequal access to education (\citeproc{ref-leeFairnessPerceptionsEducational2023}{Lee \& Stacey, 2023}). In the present paper, we argue that the perception of meritocracy in society might play a role in the justification of economic inequalities in this particular domain.

Meritocracy is defined as a distributive system where individual factors such as effort and talent are conceived as the main determinants of individual performance, usually sidelining the role of opportunities and social background. In recent years, there has been a growing interest in studying meritocratic beliefs as well as their variations according to individual and contextual variables (\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}; \citeproc{ref-mijs_belief_2022}{Mijs et al., 2022}). This line of research has highlighted how meritocratic beliefs are associated with justifying unequal access to resources and rewards in society, by legitimizing social hierarchies in which those who already have resources and advantages are more likely to succeed (\citeproc{ref-hadjar_meritokratie_2008}{Hadjar, 2008}; \citeproc{ref-mcnamee_meritocracy_2004}{McNamee \& Miller, 2004}; \citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}). Within this framework, this study focuses on the changes in the justification of inequalities in access to education in Chile, and to what extent this phenomenon is associated with the perception of meritocracy and its variations during the COVID pandemic.

The Chilean case is particularly relevant when it comes to the study of inequalities in education. The country is characterized by a high level of economic inequality associated with the privatization and commodification of several social policy domains, among them the educational system (\citeproc{ref-bellei_estudio_2013}{Bellei, 2013}; \citeproc{ref-corvalanMercadoEscolarLibertad2016}{Corvalán et al., 2016}). The neoliberal reforms implemented during and since the military dictatorship (1973-1989) have deepened educational policies based on private subsidies, achievement incentives, selection, and segregation (\citeproc{ref-joikomujicaCuasimercadoEducativoChile2019}{Joiko Mujica, 2019}). Previous research in Chile has emphasized the concept of an ``educational market'' (\citeproc{ref-corvalanMercadoEscolarLibertad2016}{Corvalán et al., 2016}), characterized by selection, competition, and the provision of goods and services associated with payment capacities. Many studies have addressed this problem from economic, political, and sociological perspectives, generating relevant evidence on their academic and social consequences (\citeproc{ref-ramosarellanoEducacionPromesaIncumplida2022}{Ramos Arellano, 2022}). Likewise, massive social mobilizations in the country (particularly in 2006, 2011, and 2018) have had educational inequalities as one of the main targets, putting pressure on governments towards a series of reforms that have advanced free higher education reforms and have diminished the role of private actors in the educational system.

This research endeavors to dissect the nuanced alterations in the justifications for educational disparities within the Chilean framework. Central to our investigation is the premise that the rationale underpinning these inequalities experiences a significant recalibration during periods of vulnerability and crises, specifically under the unprecedented strains of the global health and economic cataclysm triggered by the COVID-19 pandemic (\citeproc{ref-breznau_welfare_2021}{Breznau, 2021}; \citeproc{ref-hankivsky_introduction_2022}{Hankivsky et al., 2022}). A pertinent aspect of this discourse is articulated in the works of Mijs (\citeproc{ref-mijs_stratified_2016}{2016}, \citeproc{ref-mijs_visualizing_2018}{2018}), showing the interplay between meritocratic beliefs and the legitimization of disparities, as well as the potential erosion of meritocratic principles amidst crises. Within this framework, we point out that periods of vulnerability (such as the health and economic crisis resulting from the COVID-19 pandemic) could weaken the justification of social inequalities, a change that would be at least in part linked to a challenge of the meritocratic ideals. Furthermore, we suggest that the association between meritocratic perception and inequality justification would be weaker among lower-status groups, as they are the most affected in critical times.

@@

The justification of economic inequalities is a controversial field of study. As equality is usually considered a synonym of justice in common sense, the idea that someone could be willing to justify inequalities seems (in principle) far from what is considerably reasonable. Nevertheless, and as pointed out by Barry, the whole normative discussion in social justice can be conceived as ``the defensibility of unequal relations between people'' (\citeproc{ref-barry_theories_1989}{1989, p. 3}), whereby empirical research has consistently shown strong variations between individuals and societies regarding inequality justification (\citeproc{ref-kelley_legitimate_2021}{Kelley \& Evans, 2021}; \citeproc{ref-mijs_paradox_2021}{Mijs, 2021}; \citeproc{ref-trump_income_2018}{Trump, 2018}). Both theoretical and empirical approaches have been able to establish a clear difference between inequality and justice, which opens the question of to what extent inequality can be tolerated or even justified.

The field of studies of subjective economic inequality has been extensively developed in the social sciences, particularly in the area of social justice research (\citeproc{ref-Resh2016}{Resh \& Sabbagh, 2016}). From a sociological point of view, the works of Kluegel \& Smith (\citeproc{ref-kluegelBeliefsStratification1981}{1981}) can be considered pioneers in framing a research area comprising perceptions, beliefs, and justifications in topics such as inequality of opportunities, poverty, and distributive justice. Theoretically, the authors understand the concept of ``beliefs'' in a ``broad sense to refer to the information (veridical or non-veridical) about a phenomenon that an individual uses as a basis both for inferring other information and for action.'' (\citeproc{ref-kluegelBeliefsStratification1981}{Kluegel \& Smith, 1981, p. 30}). However, the literature has strained the concept of ``beliefs'' by more specific distinctions around the individual and structural dimensions of economic inequality. In this regard, Wegener (\citeproc{ref-wegenerIllusionDistributiveJustice1987}{1987}) has suggested that perceptions of social status and reward hierarchy are relative to the observer's status in such hierarchies (\citeproc{ref-wegenerIllusionDistributiveJustice1987}{Wegener, 1987, p. 2}). This approach established that status positions influence how people perceive things \emph{are} and, in turn, these perceptions anchor how people evaluate this inequality as just or unjust (\citeproc{ref-jassoJusticeEarningsNew1978}{Jasso, 1978}). Additionally, Janmaat (\citeproc{ref-janmaatSubjectiveInequalityReview2013}{2013}) has suggested more detailed conceptual distinctions about how people perceive economic inequality \emph{is}, how they believe or prefer it \emph{should be}, and the process of evaluating such differences as fair.

Theoretically, Liebig \& Sauer (\citeproc{ref-liebigSociologyJustice2016}{2016}) establish three fundamental distinctions in justice attitudes in the field of sociological justice research. First, \emph{order-related attitudes} refer to individual motives that explain preferences for resource allocation based on the principles of equality, need, or entitlement. Second, \emph{procedural justice attitudes} account for the degree of acceptance of unfavorable outcomes under conditions in which the procedure under which those decisions were made is considered fair (e.g., opportunities). Finally, they can be \emph{outcome-related attitudes} which refer primarily to the economic resources that a person currently obtains and the set of collective commitments associated with possessing those resources (e.g., transfers or taxes). Within this framework, we conceptualize the justification of inequality in the first area of order-related attitudes, as it expresses preferences for resource allocation that can be associated with concepts that reflect different distributive principles (\citeproc{ref-kluegel_social_1995}{Kluegel et al., 1995}; \citeproc{ref-sonhing_merit_2011}{Son Hing et al., 2011})

An important finding in the area of order-related attitudes is that beliefs about inequality are stratified by social status (\citeproc{ref-janmaatSubjectiveInequalityReview2013}{Janmaat, 2013}; \citeproc{ref-kluegelBeliefsInequalityAmericans1986}{Kluegel \& Smith, 1986}). The evidence suggests that advantaged positions in terms of socioeconomic status, such as income, education, or occupational prestige, are associated with preferences for less egalitarian class imageries (\citeproc{ref-Evans1992}{Evans et al., 1992}; \citeproc{ref-Evans2017}{Evans \& Kelley, 2017}) and greater justification of salary gaps (\citeproc{ref-hadler_why_2005}{Hadler, 2005}; \citeproc{ref-kelley_legitimate_2021}{Kelley \& Evans, 2021}). Nevertheless, and in contrast to this perspective, it has been argued that human beings are intrinsically motivated to justify the system they inhabit, even when the information available to them points in the opposite direction (\citeproc{ref-jost_psychology_2003}{Jost \& Hunyady, 2003}; \citeproc{ref-vandertoorn_twenty_2014}{van der Toorn \& Jost, 2014}). Individuals would be generally disposed to justify the economic status quo regardless of self- and group economic interest because their support partially depends on their beliefs in system-justifying ideologies (\citeproc{ref-trump_when_2020}{Trump, 2020}). As a result, what has been suggested is that low-status might perceive the status quo as just as they are more strongly attached to dominant ideologies, while high-status individuals are more prone to justify the system to a greater extent because it favors their interests and dominant position (\citeproc{ref-jost_theory_2020}{Jost, 2020}).

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[top=30mm,left=20mm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic.},
  pdfauthor={Juan Carlos Castillo; Julio Iturra; Kevin Carrasco},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Changes in the justification of educational inequalities. The
role of perceptions of inequality and meritocracy during the COVID
pandemic.}

  \author{Juan Carlos Castillo}
            \affil{%
                  Universidad de Chile
              }
          \affil{%
                  Centro de estudios del conflicto y cohesión social
                  (COES)
              }
          \affil{%
                  Núcleo milenio de desigualdades y oportunidades
                  digitales (NUDOS)
              }
        \author{Julio Iturra}
            \affil{%
                  International Graduate School of Social Sciencies
                  (BIGSSS), University of Bremen, Germany
              }
        \author{Kevin Carrasco}
            \affil{%
                  Centro de estudios del conflicto y cohesión social
                  (COES)
              }

\date{}
\begin{document}
\maketitle
\begin{abstract}
Education is considered a key tool for social mobility and equality of
opportunities. However, despite the widespread value of education,
disparities in educational outcomes persist over time. To what extent
are such educational disparities justified in society? And, what are the
main factors driving this kind of justification? This research focuses
on the changes in the justification of inequality in access to education
in Chile and their links with the perception of meritocracy. The Chilean
case offers an interesting context for this study given its high
economic inequality, deep neoliberal policies, and commodification of
social services such as education. The central argument of this article
is that the justification of inequalities weakens during periods of
vulnerability and crisis (such as the health and economic crisis
resulting from the COVID-19 pandemic), which could be linked to a
challenge of meritocratic ideals. For testing the research hypotheses we
estimate a series of longitudinal multilevel models with data from a
Chilean longitudinal panel survey (2016 -- 2022, 6 waves, N = 2,927).
Whereas we find support for the association between the perception of
meritocracy and justification of inequality, the analyses that involved
changes over time showed an increase in the justification of inequality
in education. The discussion of these results delves into the social
consequences of justifying inequality in a sensitive area as education,
as well as the persistence of meritocratic ideals despite challenging
events.

\newline
\newline

@@

This document was last modified at 2024-12-15 23:10:25

and it was last rendered at 2024-12-15 23:10:25

@@

Despite the widespread recognition of education as a key factor for
social mobility and equality of opportunities, disparities in this
domain are far from being eliminated. As the last PISA Report points
out, ``Socio-economically disadvantaged students are seven times more
likely than advantaged students to score below Level 2 in mathematics on
average across OECD countries'' (OECD 2023). Furthermore, empirical
evidence from social science research has revealed an exacerbation of
educational inequalities across various domains, such as family
socioeconomic status, gender, ethnicity, and migratory background,
particularly after COVID-19 (Darmody, Smyth, and Russell 2021).
Nevertheless, and despite the consistent empirical evidence on
educational inequalities and their consequences, research on
distributive justice indicates remarkable individual differences in the
justification of unequal access to education (Lee and Stacey 2023). In
the present paper, we argue that the perception of meritocracy in
society might play a role in the justification of economic inequalities
in this particular domain.

Meritocracy is defined as a distributive system where individual factors
such as effort and talent are conceived as the main determinants of
individual performance, usually sidelining the role of opportunities and
social background. In recent years, there has been a growing interest in
studying meritocratic beliefs as well as their variations according to
individual and contextual variables (Mijs 2021; Mijs et al. 2022). This
line of research has highlighted how meritocratic beliefs are associated
with justifying unequal access to resources and rewards in society, by
legitimizing social hierarchies in which those who already have
resources and advantages are more likely to succeed (Sandel 2020;
McNamee and Miller 2004; Hadjar 2008). Within this framework, this study
focuses on the changes in the justification of inequalities in access to
education in Chile, and to what extent this phenomenon is associated
with the perception of meritocracy and its variations during the COVID
pandemic.

The Chilean case is particularly relevant when it comes to the study of
inequalities in education. The country is characterized by a high level
of economic inequality associated with the privatization and
commodification of several social policy domains, among them the
educational system (Bellei 2013; Corvalán, Carrasco, and García-Huidobro
2016). The neoliberal reforms implemented during and since the military
dictatorship (1973-1989) have deepened educational policies based on
private subsidies, achievement incentives, selection, and segregation
(Joiko 2019). Previous research in Chile has emphasized the concept of
an ``educational market'' (Corvalán, Carrasco, and García-Huidobro
2016), characterized by selection, competition, and the provision of
goods and services associated with payment capacities. Many studies have
addressed this problem from economic, political, and sociological
perspectives, generating relevant evidence on their academic and social
consequences (Ramos 2022). Likewise, massive social mobilizations in the
country (particularly in 2006, 2011, and 2018) have had educational
inequalities as one of the main targets, putting pressure on governments
towards a series of reforms that have advanced free higher education
reforms and have diminished the role of private actors in the
educational system.

This research endeavors to dissect the nuanced alterations in the
justifications for educational disparities within the Chilean framework.
Central to our investigation is the premise that the rationale
underpinning these inequalities experiences a significant recalibration
during periods of vulnerability and crises, specifically under the
unprecedented strains of the global health and economic cataclysm
triggered by the COVID-19 pandemic (Hankivsky, Morrow, and Varcoe 2022;
Breznau 2021). A pertinent aspect of this discourse is articulated in
the works of Mijs (2016, 2018), showing the interplay between
meritocratic beliefs and the legitimization of disparities, as well as
the potential erosion of meritocratic principles amidst crises. Within
this framework, we point out that periods of vulnerability (such as the
health and economic crisis resulting from the COVID-19 pandemic) could
weaken the justification of social inequalities, a change that would be
at least in part linked to a challenge of the meritocratic ideals.
Furthermore, we suggest that the association between meritocratic
perception and inequality justification would be weaker among
lower-status groups, as they are the most affected in critical times.

@@

The justification of economic inequalities is a controversial field of
study. As equality is usually considered a synonym of justice in common
sense, the idea that someone could be willing to justify inequalities
seems (in principle) far from what is considerably reasonable.
Nevertheless, and as pointed out by Barry, the whole normative
discussion in social justice can be conceived as ``the defensibility of
unequal relations between people'' (1989, 3), whereby empirical research
has consistently shown strong variations between individuals and
societies regarding inequality justification (Trump 2018; Kelley and
Evans 2021; Mijs 2021). Both theoretical and empirical approaches have
been able to establish a clear difference between inequality and
justice, which opens the question of to what extent inequality can be
tolerated or even justified.

The field of studies of subjective economic inequality has been
extensively developed in the social sciences, particularly in the area
of social justice research (Resh and Sabbagh 2016). From a sociological
point of view, the works of Kluegel and Smith (1981) can be considered
pioneers in framing a research area comprising perceptions, beliefs, and
justifications in topics such as inequality of opportunities, poverty,
and distributive justice. Theoretically, the authors understand the
concept of ``beliefs'' in a ``broad sense to refer to the information
(veridical or non-veridical) about a phenomenon that an individual uses
as a basis both for inferring other information and for action.''
(Kluegel and Smith 1981, 30). However, the literature has strained the
concept of ``beliefs'' by more specific distinctions around the
individual and structural dimensions of economic inequality. In this
regard, Wegener (1987) has suggested that perceptions of social status
and reward hierarchy are relative to the observer's status in such
hierarchies (Wegener 1987, 2). This approach established that status
positions influence how people perceive things \emph{are} and, in turn,
these perceptions anchor how people evaluate this inequality as just or
unjust (Jasso 1978). Additionally, Janmaat (2013) has suggested more
detailed conceptual distinctions about how people perceive economic
inequality \emph{is}, how they believe or prefer it \emph{should be},
and the process of evaluating such differences as fair.

Theoretically, Liebig and Sauer (2016) establish three fundamental
distinctions in justice attitudes in the field of sociological justice
research. First, \emph{order-related attitudes} refer to individual
motives that explain preferences for resource allocation based on the
principles of equality, need, or entitlement. Second, \emph{procedural
justice attitudes} account for the degree of acceptance of unfavorable
outcomes under conditions in which the procedure under which those
decisions were made is considered fair (e.g., opportunities). Finally,
they can be \emph{outcome-related attitudes} which refer primarily to
the economic resources that a person currently obtains and the set of
collective commitments associated with possessing those resources (e.g.,
transfers or taxes). Within this framework, we conceptualize the
justification of inequality in the first area of order-related
attitudes, as it expresses preferences for resource allocation that can
be associated with concepts that reflect different distributive
principles (Kluegel, Mason, and Wegener 1995; Son Hing et al. 2011)

An important finding in the area of order-related attitudes is that
beliefs about inequality are stratified by social status (Janmaat 2013;
Kluegel and Smith 1986). The evidence suggests that advantaged positions
in terms of socioeconomic status, such as income, education, or
occupational prestige, are associated with preferences for less
egalitarian class imageries (Evans, Kelley, and Kolosi 1992; Evans and
Kelley 2017) and greater justification of salary gaps (Hadler 2005;
Kelley and Evans 2021). Nevertheless, and in contrast to this
perspective, it has been argued that human beings are intrinsically
motivated to justify the system they inhabit, even when the information
available to them points in the opposite direction (Jost and Hunyady
2003; van der Toorn and Jost 2014). Individuals would be generally
disposed to justify the economic status quo regardless of self- and
group economic interest because their support partially depends on their
beliefs in system-justifying ideologies (Trump 2020). As a result, what
has been suggested is that low-status might perceive the status quo as
just as they are more strongly attached to dominant ideologies, while
high-status individuals are more prone to justify the system to a
greater extent because it favors their interests and dominant position
(Jost 2020).

Another factor that has been linked to inequality beliefs is the
perception of economic inequality (Kelley and Evans 1993; Osberg and
Smeeding 2006; Castillo, García-Castro, and Venegas 2022). Studies in
this matter have argued that perceived economic inequality functions as
an anchor for the justifications of inequality because if these
inequalities are indeed seen as fair in their generative processes, then
they will be considered as just (Wegener 1987, 1990). Thus, both the
role of social status and economic inequality perceptions have found
support in correlational (Castillo 2012; Schröder 2017), as well as in
experimental studies that have manipulated inequality perception (Iturra
et al. 2023; Trump 2018). In addition, normative or value-driven
mechanisms can also explain to what extent people justify inequality.
For instance, studies suggest that justice ideologies such as
individualism or egalitarianism can mediate the influence of individual
status on income inequality justification (Castillo 2011). Likewise,
system justification ideologies can strengthen the influence of
socioeconomic status on economic inequality preferences (García-Sánchez
and De Carvalho 2022). Altogether, these studies have shown that not
only socio-structural factors influence what people believe regarding
the distribution of economic resources, but their normative stances and
perceptions can also determine preferences for economic inequality in
diverse domains.
